Distributed Computing for the Common Man

### 2010-07-01 doc/ml-workshop-2010/main.tex
*** empty log message ***

In addition to the above, distributed computing environments also
implement mechanisms for fault tolerance, efficient storage, and
distribution of tasks. This is required to handle network failures
that may occur, as well as to optimize the usage of machines in the
network. Another concern of importance is the transmission of messages
over the network. This requires efficient 
\emph{marshaling} of data, that is encoding and decoding of data 
for transmission over different computing environments.  It is desirable to
maintain architecture independence while transmitting marshalled data,
as machines in a distributed computing environment often run on
different hardware architectures and make use of varying software
platforms. For example, machine word size or endianness may be different
across machines on the network.

\paragraph{A Functional Programming Approach.}
Our work was initially inspired by Google's
MapReduce\footnote{Ironically, Google's approach itself was inspired
  by functional programming primitives.}. However, our functional
programming environment allows us to be more generic. 
The main idea behind our approach is that
workers may implement any polymorphic function:
\begin{ocaml}
  worker: 'a -> 'b
\end{ocaml}
where \of{'a} denotes the type of tasks and \of{'b} the type of results.
Then the master is given a list of initial tasks, together with a
function to handle the results:
\begin{ocaml}
  master: ('a -> 'b -> 'a list) -> 'a list -> unit
\end{ocaml}
The function passed to the master is applied whenever a result is
available and may in turn generate new tasks (hence the return type
\of{'a list}).  The master is executed as long as there are pending
tasks.

Our library makes use of \Ocaml's marshaling capabilities as much as
possible. Whenever master and worker executables are exactly the same,
we can marshal polymorphic values and closures. However, it is not
always possible to have master and workers running the same
executable. In this case, we cannot marshal closures anymore but we
can still marshal polymorphic values as long as the same version of
\Ocaml\ is used to compile master and workers. When different versions
of \Ocaml\ are used, we can no longer marshal values but we can still
transmit strings between master and workers. Our library adapts to all
these situations, by providing several APIs.

\medskip
The paper is organized as follows. In Section~\ref{sec:API}, we
describe the main idea behind the generic API, and introduce
\of{master} and \of{worker} programs. Section~\ref{sec:example}
illustrates the API usage with an example.
In Section~\ref{sec:derived}, we
derive some interesting functions from the generic
API. Section~\ref{sec:implem} delves into the implementation details
of our library, while Section~\ref{sec:experiments} illustrates the
potential of the presented library through experimental evaluation. We
compare our approach with relevant related work in
Section~\ref{sec:future} and outline our future work.

Our library, \functory, is implemented in \Ocaml\ and is available
from \url{http://www.lri.fr/~filliatr/functory/}.

TODO : refer to the technical report available on-line

@@

Our work was initially inspired by Google's
MapReduce\footnote{Ironically, Google's approach itself was inspired
  by functional programming primitives.}, from where we borrow some 
terminology:
\begin{itemize}
\item A notion of \emph{tasks} which denote atomic computations
  to be performed in a distributed manner; 
\item A set of processes (possibly executing on remote machines)
  called \emph{workers} that perform
  the tasks, producing results;
\item A single process called a \emph{master} which is in charge
  of distributing the tasks among the workers and managing results
  produced by the workers.
\end{itemize}
The main function of our API combines these three notions of tasks,
master and workers into a single higher-order polymorphic function
\of{compute}:
\begin{ocaml}
  val compute : 
    worker:('a -> 'b) -> 
    master:('a * 'c -> 'b -> ('a * 'c) list) -> 
    ('a * 'c) list -> unit
\end{ocaml}
Tasks are pairs, of type \of{'a * 'c}, where the first component is
passed to the worker and the second component is local to the master.
The \ocaml{worker} function should be pure and is executed in parallel
in all worker processes. The function \ocaml{master}, on the
contrary, can be impure and is only executed in the master process.
The \ocaml{master} function typically stores results 
in some internal data structure.

@@

  \begin{ocaml}
    val map : 
      f:('a -> 'b) -> 'a list -> 'b list 
    val map_local_fold : 
      map:('a -> 'b) -> fold:('c -> 'b -> 'c) -> 
      'c -> 'a list -> 'c 
    val map_remote_fold : 
      map:('a -> 'b) -> fold:('c -> 'b -> 'c) -> 
      'c -> 'a list -> 'c 
    val map_fold_ac : 
      map:('a -> 'b) -> fold:('b -> 'b -> 'b) -> 
      'b -> 'a list -> 'b 
    val map_fold_a : 
      map:('a -> 'b) -> fold:('b -> 'b -> 'b) -> 
      'b -> 'a list -> 'b
  \end{ocaml}

@@

Let us assume we need to compute a sum such as
\begin{displaymath}
  s(n) = \sum_{i=0}^{n}f(i)
\end{displaymath}
where $f$ is a pure and computationally expensive function.
We start with the sequential implementation provided by the library:
\begin{ocaml}
  open Sequential
\end{ocaml}
The idea is to test our code on small values of $n$ before
parallelizing the computation for larger values.
Let us assume that $f$ is implemented as a top-level \of{worker} function:
\begin{ocaml}
  let worker i = ...
\end{ocaml}
and that $n$ is obtained from the command line:
\begin{ocaml}
  let n = int_of_string Sys.argv.(1)
\end{ocaml}
The list of tasks is simply the list of integers $0,1,\dots,n$,
computed as follows:
\begin{ocaml}
  let tasks = 
    let l = ref [] in 
    for i = 0 to n do l := (i, ()) :: !l done; 
    !l
\end{ocaml}
Each task is actually a pair, where the second component contains
some information which is kept local to the master process.
It is irrelevant here and thus the value \of{()} of type \of{unit} is used.
The current sum is kept into a global reference \of{s}:
\begin{ocaml}
  let s = ref 0
\end{ocaml}
The master function receives a task \of{(i,())} together with its
result \of{fi} and simply adds \of{fi} to \of{s}. There is no new task
generated, hence the returned empty list.
\begin{ocaml}
  let master (i,()) fi = s := !s + fi; []
\end{ocaml}
Finally, the whole computation takes place when we call the
\of{compute} function:
\begin{ocaml}
  let () = compute ~worker ~master tasks
\end{ocaml}

Now, let us assume we want to make use of a 4-core machine to compute
$s(n)$. It is as easy as just replacing the very first line with
\begin{ocaml}
  open Cores
  let () = set_number_of_cores 4
\end{ocaml}
while the rest of the code remains unchanged.

Later, we may want to use a network of machines instead, for example
two machines called \texttt{orcus} and \texttt{belzebuth} with 4 and 8
cores respectively. Similar to the above, we just replace the first
lines with
\begin{ocaml}
  open Network
  let () = declare_workers ~n:4 "orcus"
  let () = declare_workers ~n:8 "belzebuth"
  open Same
\end{ocaml}
Here, \of{Same} is a module which is to be used when master and worker
are running the same executable. It provides a \of{compute} function
which has the same signature as in modules \of{Sequential} and
\of{Cores}, so that the rest of the code remains unchanged. Master and
worker processes are distinguished at run-time using an environment
variable \texttt{WORKER} which is set/unset.

If we need to write two different programs for master and worker, for
reason of binary incompatibility or any other reason, the library API
is providing functions to do that. If master and worker are still
compiled with the same version of \Ocaml, we use the \of{Poly} module
which provides a polymorphic API. Let us start with the worker program.
It is as simple as
\begin{ocaml}
  open Poly
  let worker i = ...
  let () = Worker.compute worker ()
\end{ocaml}
The \of{Worker.compute} function enters a loop which waits for
tasks sent by the master and returns results computed using \of{worker}.
The master program is almost the same as before. First, we replace
module \of{Same} with module \of{Poly}:
\begin{ocaml}
  open Network
  let () = declare_workers ~n:4 "orcus"
  let () = declare_workers ~n:8 "belzebuth"
  open Poly
\end{ocaml}
Tasks and \of{master} function are unchanged:
\begin{ocaml}
  let tasks = ...
  let s = ref 0
  let master (i,()) fi = ...
\end{ocaml}
Finally, we start the computation with \of{Master.compute}, which does
not have a \of{worker} parameter anymore:
\begin{ocaml}
  let () = Master.compute ~master tasks
\end{ocaml}

When master and worker programs are compiled with different
versions of the \Ocaml\ compiler, our library still provides
a monomorphic API over strings. As a consequence, we need to convert
integers to and from strings in both master and worker.
The modified worker program is as follows:
\begin{ocaml}
  open Mono
  let worker i = ...
  let worker_string i = string_of_int (worker (int_of_string i))
  let () = Worker.compute worker_string ()
\end{ocaml}
The master program is modified in a similar way.
We simply replace \of{Poly} with \of{Mono} and encode/decode integers
as strings, as follows:
\begin{ocaml}
  let tasks = 
    let l = ref [] in 
    for i = 0 to n do l := (string_of_int i, ()) :: !l done; 
    !l
  let master (i,()) fi = s := !s + int_of_string fi; []
\end{ocaml}

>>>>

% In addition to the above, distributed computing environments also
% implement mechanisms for fault tolerance, efficient storage, and
% distribution of tasks. This is required to handle network failures
% that may occur, as well as to optimize the usage of machines in the
% network. Another concern of importance is the transmission of messages
% over the network. This requires efficient 
% \emph{marshaling} of data, that is encoding and decoding of data 
% for transmission over different computing environments.  It is desirable to
% maintain architecture independence while transmitting marshalled data,
% as machines in a distributed computing environment often run on
% different hardware architectures and make use of varying software
% platforms. For example, machine word size or endianness may be different
% across machines on the network.

% \paragraph{A Functional Programming Approach.}
% Our work was initially inspired by Google's
% MapReduce\footnote{Ironically, Google's approach itself was inspired
%   by functional programming primitives.}. However, our functional
% programming environment allows us to be more generic. 
% The main idea behind our approach is that
% workers may implement any polymorphic function:
% \begin{ocaml}
%   worker: 'a -> 'b
% \end{ocaml}
% where \of{'a} denotes the type of tasks and \of{'b} the type of results.
% Then the master is given a list of initial tasks, together with a
% function to handle the results:
% \begin{ocaml}
%   master: ('a -> 'b -> 'a list) -> 'a list -> unit
% \end{ocaml}
% The function passed to the master is applied whenever a result is
% available and may in turn generate new tasks (hence the return type
% \of{'a list}).  The master is executed as long as there are pending
% tasks.

% Our library makes use of \Ocaml's marshaling capabilities as much as
% possible. Whenever master and worker executables are exactly the same,
% we can marshal polymorphic values and closures. However, it is not
% always possible to have master and workers running the same
% executable. In this case, we cannot marshal closures anymore but we
% can still marshal polymorphic values as long as the same version of
% \Ocaml\ is used to compile master and workers. When different versions
% of \Ocaml\ are used, we can no longer marshal values but we can still
% transmit strings between master and workers. Our library adapts to all
% these situations, by providing several APIs.

% \medskip
% The paper is organized as follows. In Section~\ref{sec:API}, we
% describe the main idea behind the generic API, and introduce
% \of{master} and \of{worker} programs. Section~\ref{sec:example}
% illustrates the API usage with an example.
% In Section~\ref{sec:derived}, we
% derive some interesting functions from the generic
% API. Section~\ref{sec:implem} delves into the implementation details
% of our library, while Section~\ref{sec:experiments} illustrates the
% potential of the presented library through experimental evaluation. We
% compare our approach with relevant related work in
% Section~\ref{sec:future} and outline our future work.

Our work was initially inspired by Google's
MapReduce\footnote{Ironically, Google's approach itself was inspired
  by functional programming primitives.}, from where we borrow some 
terminology:
\begin{itemize}
\item A notion of \emph{tasks} which denote atomic computations
  to be performed in a distributed manner; 
\item A set of processes (possibly executing on remote machines)
  called \emph{workers} that perform
  the tasks, producing results;
\item A single process called a \emph{master} which is in charge
  of distributing the tasks among the workers and managing results
  produced by the workers.
\end{itemize}

Our library, \functory, is implemented in \Ocaml.
It makes use of \Ocaml's marshaling capabilities as much as
possible. Whenever master and worker executables are exactly the same,
we can marshal polymorphic values and closures. However, it is not
always possible to have master and workers running the same
executable. In this case, we cannot marshal closures anymore but we
can still marshal polymorphic values as long as the same version of
\Ocaml\ is used to compile master and workers. When different versions
of \Ocaml\ are used, we can no longer marshal values but we can still
transmit strings between master and workers. Our library adapts to all
these situations, by providing several APIs.

Our library is available
at \url{http://www.lri.fr/~filliatr/functory/}.
A report is also available there, which provides more
technical details.

@@

The main function of our API combines these three notions of tasks,
master and workers into a single higher-order polymorphic function
\of{compute}:
\begin{ocaml}
  val compute : 
    worker:('a -> 'b) -> 
    master:('a * 'c -> 'b -> ('a * 'c) list) -> 
    ('a * 'c) list -> unit
\end{ocaml}
Tasks are pairs, of type \of{'a * 'c}, where the first component is
passed to the worker and the second component is local to the master.
The \ocaml{worker} function should be pure and is executed in parallel
in all worker processes. The function \ocaml{master}, on the
contrary, can be impure and is only executed in the master process.
The \ocaml{master} function typically stores results 
in some internal data structure.

@@

In addition to this low-level API, our library also provides more
traditional \of{map} and \of{fold} higher-order functions with
distributed computation. For instance, 
\begin{ocaml}
  val map_local_fold : 
    map:('a -> 'b) -> fold:('c -> 'b -> 'c) -> 
    'c -> 'a list -> 'c 
\end{ocaml}
applies \of{map} to each element of the list in a distributed way and
folds the results locally to the master.

@@

Let us assume we need to compute a sum such as
\begin{displaymath}
  s(n) = \sum_{i=0}^{n}f(i)
\end{displaymath}
where $f$ is a pure and computationally expensive function.
Using the sequential implementation and the low-level \of{compute}
function, the code looks like:
\begin{ocaml}
1.  open Sequential
2.  let f i = ...
3.  let n = int_of_string Sys.argv.(1)
4.  let tasks = ... (* the lists [0; 1; ...; n] *)
5.  let s = ref 0
6.  let master (i,()) fi = s := !s + fi; []
7.  let () = compute ~worker:f ~master tasks
\end{ocaml}
To switch to the multi-core implementation, say with 4 cores, we
simply replace line 1 with
\begin{ocaml}
1a.  open Cores
1b.  let () = set_number_of_cores 4
\end{ocaml}
Similarly, in order to use a network of two machines \texttt{m1} and
\texttt{m2}, with 4 and 8 cores respectively, line 1 is replaced with
\begin{ocaml}
1a.  open Network
1b.  let () = declare_workers ~n:4 "m1"
1c.  let () = declare_workers ~n:8 "m2"
\end{ocaml}

We could simplify the code even further using the derived API, for
example by replacing lines 5-7 with
\begin{ocaml}
5.  let s = map_local_fold ~map:f ~fold:(+) 0 tasks
\end{ocaml}
